Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19

### 2023-04-11 manuscript/manuscript.qmd
Suparna edits

Do emergency provisions during crises do more harm or good when it comes to human rights? Combating a deadly virus like SARS-CoV-2 required extraordinary public health measures. In an effort to fight the COVID-19 pandemic, governments across the world imposed restrictions on the freedom of movement, association, assembly and privacy. While countries were ostensibly free to instate these policies within their own borders, the imposition of these restrictions raised important questions about states' commitments to international human rights law. Some human rights treaties recognize states' needs for emergency measures during public health concerns and contain escape clauses that establish safeguards to prevent permanent erosion of these rights during emergencies. The temporary suspension of state compliance with its international treaty obligations occurs through a legal action known as derogation. However, the potential abuse of emergency measures during derogations raises concerns about the long-lasting impact of emergency measures on human rights. Under emergency decrees, states often impose restrictions that are temporary in theory but difficult to roll back or contain in practice [@Stasavage2020; @LuhrmannRooney2021].

Given recent medical advancements and dwindling concerns about the pandemic, the Biden administration announced that it would end the public health emergency in the US on May 11, 2023, ending measures instituted at the beginning of the pandemic—a move that has occurred in many other countries as well. With the lifting of these measures, we can assess important descriptive and analytical questions raised regarding states' respect for international law and human rights during this emergency period. What did derogations mean for compliance? Did states deviate from treaty provisions when they said they would? Are derogating and non-derogating states different in terms of how they implemented emergency measures during COVID-19? Were derogating states more likely to exploit the provisions and to abuse human rights during COVID-19? These questions are important for the study of international law, compliance, and times of crises.

The potential violations of human rights during this period also raises questions about factors that can mitigate the adverse impacts of emergency measures. Existing literature on emergencies such as wars and natural disasters shows that civil society organizations (CSOs) play a pivotal role during these periods [@pellingdill; @cash2013]. They provide services and resources, and hold governments accountable to ensure inclusive policies during these emergencies. Human rights compliance literature also generally finds that NGOs and civil society are key actors in facilitating treaty obligation actualization [e.g., @KeckSikkink:1998; @Merry:2009]. Can these groups also prevent or reduce the likelihood of human rights violations during public health emergencies? Does a robust civil society mitigate the adverse effects arising from international treaty derogations during the pandemic?

@@

Using data from the Varieties of Democracy PanDem dataset, the Oxford Covid-19 Government Response Tracker, and United Nations treaty derogation data, we first assess the variation in human rights responses among governments during the pandemic. We then analyze the effect of derogating (or not) on human rights practices during the pandemic and assess the role of civil society in mitigating the effects of rights-violating policies. We find that states that derogated and specified time limits to the emergency provisions generally stuck to the time limits and adhered to standards required during derogation by avoiding discriminatory policies and not violating non-derogable rights. On the other hand, states that did not derogate from their international human rights obligations and still imposed emergency measures were more likely impose discriminatory policies, violate non-derogable rights (the right to life, freedom from torture, and freedom from slavery), and impose emergency measures without time limits.

We also find that civil society can play an important mitigating role in curbing discriminatory policies. States with low levels of civil society repression were the least likely to abuse human rights when enforcing emergency policies. The likelihood of discriminatory policies increased as civil society repression worsened, but derogations appear to protect against worsening abuses—even in countries with severe civil society repression, those that derogated had a nearly 100% probability of no discriminatory policies.

This research note makes two contributions. First, while there is a significant literature on compliance with international law, especially human rights treaties [e.g. @Simmons2009; @Comstock2021], and the flexibility tools such as reservations, understandings, and declarations (RUDs) that states invoke when they join or withdraw from such treaties [@Neumayer:2007; @Helfer:2002; @Helfer:2005], we know little about states' compliance with flexibility mechanisms such as derogations. Compliance with derogations may not necessarily be the same as compliance with the treaty and this research note provides a systematic understanding of the types of states that are likely to comply with these escape clauses. The pandemic further provides unique insight into derogation behavior because numerous states derogated under the same global conditions of a public health emergency, rather than domestic conflict or natural disaster typical of treaty derogations, making it easier to compare the long-term human rights effects of these derogations cross nationally.

Second, there is a robust literature on compliance with international law and how it can lead to improvements in domestic human rights conditions [@Simmons2009; @Merry:2009]. While derogations are intended to make sure government responses are temporary, proportional and non-discriminatory, some legal scholars warn of the possibility that some treaty actions removing legal obligations may actually contribute to more violations and abuse [e.g. @Lebret:2020; @helfer2021] as part of a broader trend of targeting groups during times of crises [@Grasse2021]. We find that overall, derogating states do not increase abuse and violations. This finding speaks to international legal compliance scholarship and supports the idea that states are being sincere about their intentions and not, byinlarge, using actions like derogation to cover abusive behavior.

The rest of the paper proceeds as follows. First, we explore the purpose of derogations during the pandemic. In doing so, we focus specifically on derogations to the International Covenant on Civil and Political Rights (ICCPR)—arguably the most significant and extensive treaty in the international human rights regime. Following this, we introduce our expectations regarding how formally derogating from the treaty (or not) may impact a state's human rights practices during the pandemic. We then assess states' behavior towards civil society and these organizations' potential role in mitigating the adverse effects of emergency measures. We conclude with the implications of our findings for the study of international law, times of crises, and the role of civil society.

@@

Derogations in international law authorize states to temporarily suspend their international treaty commitments during times of crises, providing them with flexibility in responding to events. Times of crises can include wars, natural disasters, internal political unrest such as terrorism, and public health emergencies. Derogations afford states the opportunity to adjust to these changing conditions [@HBHelferFariss]. When states derogate from their international legal obligations,the derogation provides vital information to both international and domestic monitoring bodies, interest groups, and advocates about which rights are suspended, for how long, and the reasoning behind these suspensions. This information allows these actors—at least in principle—to challenge measures that are excessive, vague, or outlast the intended time frame of their implementation [@helfer2021].

While legal critical assessments of derogations find importance in their use [@richardson2020], others argue that, "When a state does not bother with derogation provisions in times of emergencies where human rights are suspended, then it is a situation of unbridled state power.” [@burchill2005, 97] An abuse of state power can also potentially lead to the violation of non-derogable rights—rights such as right to life, freedom from torture, freedom from slavery, freedom of thought and conscience, among others, that are viewed as essential rights that cannot be suspended even during emergencies. Research finds results that support this notion—there is no meaningful variation in states' respect for derogable or non-derogable rights during officially declared states of emergencies [@richardsclay2012; @neumayer2013]. Even states that officially derogate from their international legal obligations are more likely to violate physical integrity rights, which are non-derogable rights [@neumayer2013; @comstock2019].

These findings suggest that states may not be discerning the significance of derogable v/s. non-derogable rights status during emergencies. This also raises the question of the extent to which this distinction is a valuable one during the pandemic. Most derogations to the ICCPR have been for internal conflict or natural disaster - is a public health emergency different from other states of emergency? Overall, there is a general expectation that following derogations, we may observe some decline in respect for human rights. However, previous studies have not explored the relationship between derogations and public health emergencies.

@@

A first question of interest before we examine the potential impact of derogations is what states are more likely to submit them in the first place. Though not much research has focused on derogations in political science, and research that does exist has mostly focused on the determinants of submitting derogations to the UN rather than implications of when they are in effect. This literature tells us that states that are more democratic are more likely to issue derogations and other post-commitment actions such as reservations, understandings, and declarations [@Simmons2009; @comstock2019] and that most states facing an emergency do not, in fact, issue derogations [@HBHelferFariss]. The expectation is that states with strong democratic institutions need to convince domestic audiences of the legality and legitimacy of suspending rights during emergencies. Additionally, democratic states understand international law and international legal mechanisms to be more legitimate and recognize their importance more so than other states. Less democratic states, even though at times may be more likely to violate or evade international human rights law, are less likely to see nuanced pieces of treaty law as important.

@@

To legally derogate, government must first officially declare a state of emergency (Article 4). Derogations must fulfill three criteria: they should be temporary in application, proportional to a crisis' actual threat and non-discriminatory. However, it is worth noting that states cannot derogate from the entirety of the ICCPR. Seven rights enshrined in the ICCPR are non-derogable: the right to life, the prohibition of torture or cruel, inhumane punishment,the prohibition of slavery, the right to recognition as a person before the law, and the protection of freedom of thought, conscience and religion. The international legal community viewed these rights as so fundamental and important that no type of crisis could justify their suspension. Other types of rights are seen as able to be suspended during crises. Several ICCPR Articles provide that public health needs can justify limitations on or derogations from certain rights [@richardson2020, 112]: freedom of movement (Article 12), freedom of expression (Article 19), and the rights to peaceful assembly and freedom of association (Article 22).

Recognizing that public health exigencies would require states to curb civil and political rights, the UN Human Rights Office of the High Commissioner (OHCHR) published guidance stating that any emergency measures must meet specific criteria of legality, necessity, proportionality, and non-discrimination. However, not every country formally derogated—a number of countries implemented emergency measures without formally derogating from ICCPR. Over 100 countries issued emergency declarations while fewer than 30 submitted COVID-19 related derogations [@icnl-covid]. These COVID-19 emergency declarations may have longer term impacts beyond temporary human rights violations [@BrownBrechenmacherCarothers:2020]. Below, we first theorize how derogations can protect against rights violations and the excesses of emergency measures, followed by the role of civil society in mitigating these violations.

# The impact of derogations on human rights outcomes

@@

During the pandemic, as @helfer2021 notes, states have had “sharply divergent responses” to international human rights law even when they are parties to the same treaty and confronting similar threats (21). If states determine that emergency measures are justified without regard for international law, the processes for restoring rights at the end of the emergency may not be followed. As such, many states used the pandemic an as excuse to centralize authority and implemented measures that didn’t meet international legal standards. Many states imposed restrictions on travel or movement by minority or opposition groups, limiting public assembly and democratic protests, and deployed surveillance systems that impeded privacy. For instance, in Australia, indigenous and migrant communities were disproportionately targeted in the enforcement of movement restrictions [@ICRC2021]. In Bulgaria, checkpoint controls went into effect in neighborhoods largely composed of the Roma community [@bulgaria2020]. In India, the police used a ban on public gatherings as an excuse to get rid of a months-long sit-in protesting the country's new citizenship law seen as discriminatory towards Muslims [@bbc2020].

We expect that states that did not notify the ICCPR of impending rights suspensions and submit derogations were less likely to take the criteria for derogations—necessity, proportionality, non-discrimination and time delimited seriously, and therefore more likely to impose measures that violated these criteria. If states determined that emergency measures are justified without consideration of international law, we expect that there will be more violations.

@@

## Civil society and human rights

Given the divergent state responses during the pandemic, differing responses can have lasting impacts on human rights trends. In the absence of adequate review from the international community during the pandemic, certain domestic-level factors may be necessary to mitigate the adverse effects of emergency measures. While both the theoretical and empirical literature has established that states may violate both derogable and non-derogable rights during emergencies, we have comparatively little evidence regarding factors that may safeguard against these rights violations during periods of derogation.

There is a robust literature on how civil society is important for improving human rights outcomes [@KeckSikkink:1998; @RisseRoppSikkink; @landman2005; @HBTsutsui; @neumayer2005; @Jetschke2011]. Previous literature finds that an effective civil society response can mitigate the impact of natural disasters and crises and even increase political engagement in the long-run [@fair2017]. As such, we expect that a strong and vibrant civil society may help mitigate the adverse impact arising from derogations and emergency measures imposed by governments during the pandemic. Civil society organizations broaden the range of voices in the policymaking process and help amplify the concerns and interests of citizens [@murdie2014]. Civic initiatives have not only played a major role in countering the pandemic's severity, but have also pressured authorities into implementing better policies to deal with the impact of economic shutdowns, product shortages and spikes in unemployment. For instance, in Brazil, a broad coalition of CSOs lobbied the government to implement a relief fund, successfully securing a monthly stipend three times the amount initially suggested by the government [@carnegie2021, 5].

Government attitudes towards civil society organizations—whether collaboration, accommodation or repression—may also affect their respect (or lack thereof) for human rights during the pandemic. Since March 2020, many advocacy and rights-based organizations have highlighted how the pandemic disproportionately affects vulnerable and marginalized communities. These organizations also monitor and document human rights violations by the state, which can both provide information about a state's level of compliance with domestic and international human rights obligations as well can be used as evidence in court.

However, over the past three decades, more than 130 states have repressed civil society organizations [@chaudhry2022]. These trends have been further exacerbated during the pandemic. As a result, a number of advocacy and rights-based organizations have either shut down or significantly downsized their capabilities, hampering their ability to document and monitor rights violations by the state and its agents. Without such monitoring, states may be more likely violate non-derogable rights during the pandemic, as states know that organizations will be less likely to collect evidence, write briefs, provide legal aid to victims and give expert testimony at trials. As such, we expect that states will have greater leverage to impact non-derogable rights if there is insufficient evidence that can be presented to international monitoring institutions or regional courts such as the European Court of Human Rights.

> **Hypothesis 2~a~**: If states are repressive towards civil society organizations, then they are more likely to violate non-derogable rights.

Given reduced monitoring by rights-based organizations, states may also have greater leeway to pass discriminatory policy measures and engage in more abusive enforcement of these measures. This is particularly the case for state actions towards refugees and asylum-seekers. For instance, Greece has increasingly criminalized humanitarian assistance and in 2019, introduced burdensome rules for the registration of nongovernmental organizations working on migration [@Cosse:2022]. Once the pandemic began, with decreased oversight from CSOs owing to the above rules, Greek officials intercepted and turned back boats filled with asylum-seekers and also summarily returned asylum-seekers at the land border with Turkey. In doing so, they both violated the non-derogable principle of non-refoulement, as well as Article 7 of ICCPR is the person sent back to another state where they are in danger of ill-treatment or torture [@HathawayStevensLim:2020].

> **Hypothesis 2~b~**: If states are repressive towards civil society organizations, then they are more likely to use discriminatory policy measures during the pandemic.
> 
> **Hypothesis 2~c~**: If states are repressive towards civil society organizations, then they are more likely to engage in abusive enforcement of measures instated to deal with the pandemic.

Finally, this trend of state repression of CSOs and NGOs may also impact state behavior towards the media. States typically repress CSOs because they make democratic participation, particularly protests, more likely, highlight rights violations, and challenge economic policies—all of which may be seen as even more threatening during the pandemic [@MurdieBhasin2011; @BBCM2014; @KeckSikkink:1998; @RisseRoppSikkink; @Boulding2014; @chaudhry2022].

With many governments trying to centralize authority and decisionmaking, a free and independent media may pose challenges similar to those posed by CSOs [@Themudo:2013]. As such, if states are engaging in repression of CSOs during the pandemic, they may be more likely to repress the media as well. For instance, over the last five years, Hungary has increasingly cracked down on CSOs. During the pandemic, this extended towards the media as well as the country passed a new law which sought to punish anyone who published 'false information' on the pandemic with up to five years in jail [@brechenmacheretal2020, 3]. Similarly, Zimbabwe criminalized false reporting on the government's response to the pandemic and subsequently, used this law to crack down on media freedom. On the other hand, a vibrant civil society may help protect against limitations on media freedom. For instance, in Tunisia, the parliament withdrew a draft law restricting media freedom after it faced criticism from CSOs [@brechenmacheretal2020, 3].

> **Hypothesis 2~d~**: If states are repressive towards civil society organizations, then they are more likely to impose limits on media freedom during the pandemic.

@@

While some studies on repression, human rights, and the pandemic have focused within single country cases or across a region, we examine a global database to more broadly examine the relationship between treaty derogations, civil society, and state actions during the pandemic. We explore our hypotheses using data on COVID-19 derogations, state actions, health trends, and civil society measures that we collected from several different sources (see @tbl-data-sources), including start and end dates of ICCPR derogations [@UnitedNations:2023], formal government responses from the Oxford Covid-19 Government Response Tracker [@HaleAngristGoldszmidt:2021], indicators of the severity of government responses from the Varieties of Democracy project [@vdem-v12; @vdem-pandem], and COVID case and death counts from the World Health Organization's COVID-19 dashboard [@WHO:2023]. We include data from `r info$n_countries` countries during the first 15 months (or `r info$n_weeks` weeks) of the pandemic, from March 11, 2020 to June 30, 2021. While the Oxford Government Response Tracker provides daily data on a variety of government actions, submitting formal treaty responses to the ICCPR is less likely to be as immediate of an action as implementing domestic emergency actions like stay at home orders. Accordingly, we collapse this daily data into weekly totals, resulting in a panel with `r info$n_rows` country-week observations.

@@

# Discussion and conclusion

The first few months of the pandemic raised numerous concerns that the pandemic would contribute to violations of democratic standards and human rights in a number of countries and would reinforce the powers of repressive governments worldwide. Derogations are an international legal means to communicate to both the UN and global community that certain rights will be suspended during times of crisis. This research note set out to examine state responses to their international human rights treaty obligations, particularly the ICCPR, during the COVID-19 pandemic and any subsequent impact on human rights violations.

Derogations are meant to provide states with flexibility during crises and to encourage them to sign and ratify treaties without concern about the inability to adapt to times of emergency. International legal scholarship has been divided over expectations of derogations during the pandemic. @Lebret:2020 argued that derogations are the best legal mechanism for states to notify the global community of their inability to comply with human rights standards during the COVID-19 pandemic. However, @richardson2020 see derogations as a less clear path for states during COVID-19, describing the actions, their conditions, and human rights as a "tangled morass" (p. 106).

With the caveat that our results are by no means causal—we have not attempted to use any formal strategy to identify an unconfounded causal effect, and our findings are merely descriptive—we find that we find that there is good news for compliance and derogations. Derogators did not systematically use this legal "escape clause" opportunistically—derogators used the process correctly. Most derogations included time limitations, states tended to adhere to the time limits, and there was not an overall abuse of the derogation time to abuse, repress, and discriminate. Though not many states submitted COVID-19 related derogations, it seems that the ones that did used derogations as intended by international law. Once the time of derogation ended, states readjusted to typical compliance once more. States, therefore complied with the intent of derogations and the purpose of the treaty. This is an important finding for international legal behavior and compliance studies. Even in a time of crisis, overall, states considered it important to follow through with the processes and expectations of international human rights law bureaucracy.

We suggest that future research look to differences in levels within states during crisis. There may be key differences in enforcement between local level bureaucrats and higher level elites. While higher level diplomats submit derogations and may be more cautious about international law, local level enforcement may not be as concerned about international legal aspects such as time limits and derogable vs. non-derogable rights. This distinction may explain some variation in policy and discrimination practices during the pandemic. We also recommend that future research compare COVID-19 area derogation compliance with that in other crises. This note contributed to the study of derogations by building on the analysis between derogation submission and outcome. States did end up taking the restriction measures they signaled they would via the derogations. Finally, the relationship and interaction between civil society, human rights, and international law remains an important and ripe field for future research.

>>>>

What is the impact of escape clauses or emergency provisions in international law on human rights? Combating a deadly virus like SARS-CoV-2 required extraordinary public health measures. In an effort to fight the COVID-19 pandemic, governments across the world imposed restrictions on the freedom of movement, association, assembly and privacy. While countries were ostensibly free to instate these policies within their own borders, the imposition of these restrictions raised important questions about states' commitments to international human rights law. Some human rights treaties recognize states' needs for emergency measures during public health concerns and contain escape clauses that establish safeguards to prevent the permanent erosion of these rights during emergencies. The temporary suspension of state compliance with its international treaty obligations occurs through a legal action known as derogation. However, the potential abuse of emergency measures during derogations raises concerns about their impact on human rights. Under emergency decrees, states often impose restrictions that are temporary in theory but difficult to roll back or contain in practice [@Stasavage2020; @LuhrmannRooney2021].

Given recent medical advancements and dwindling concerns about the pandemic, the Biden administration announced that it would end the public health emergency in the US on May 11, 2023, lifting measures instituted at the beginning of the pandemic—a move that has occurred in many other countries as well. With this move, we can assess important descriptive and analytical questions raised regarding states' respect for international law and human rights during this emergency period. What did derogations mean for compliance with international law? Did derogating and non-derogating states differ in terms of how they implemented emergency measures during COVID-19? Were derogating states more likely to exploit emergency provisions and to abuse human rights during COVID-19? These questions are important for the study of international law, compliance, and human rights during times of crises.

@@

Using data from the Varieties of Democracy PanDem dataset, the Oxford Covid-19 Government Response Tracker, and United Nations treaty derogation data, we first assess the variation among derogating and non-derogating states on the implementation of emergency measures. We then analyze the effect of derogating (or not) on human rights practices during the pandemic. We find that states that derogated and specified time limits to the emergency provisions generally stuck to the time limits and adhered to standards required during derogation by avoiding discriminatory policies and not violating non-derogable rights. On the other hand, states that did not derogate from their international human rights obligations and still imposed emergency measures were more likely to impose discriminatory policies, violate non-derogable rights (the right to life, freedom from torture, and freedom from slavery), and impose emergency measures without time limits.

This research note makes two contributions. First, while there is a significant literature on compliance with international law, especially human rights treaties [e.g. @Simmons2009; @Comstock2021], and the flexibility tools such as reservations, understandings, and declarations (RUDs) that states invoke when they join or withdraw from such treaties [@Neumayer:2007; @Helfer:2002; @Helfer:2005], we know little about states' compliance with flexibility mechanisms such as derogations. Compliance with derogations may not necessarily be the same as compliance with the treaty and this research note provides a systematic understanding of the types of states that are likely to comply with these escape clauses. The pandemic further provides unique insight into derogation behavior because numerous states derogated under the same global conditions of a public health emergency, rather than domestic conflict or natural disaster typical of treaty derogations, making it easier to compare state behavior during these derogations cross nationally. In making this comparison, this research note makes an important descriptive contribution showing how the implementation of policy measures differed across derogating and non-derogating states during the pandemic.

Second, there is a robust literature on compliance with international law and how it can lead to improvements in domestic human rights conditions [@Simmons2009; @Merry:2009]. While derogations are intended to make sure government responses are temporary, proportional and non-discriminatory, some legal scholars warn of the possibility that some treaty actions removing legal obligations may actually contribute to more violations and abuse [e.g. @Lebret:2020; @helfer2021] as part of a broader trend of targeting groups during times of crises [@Grasse2021]. We find that overall, derogating states during the pandemic do not increase abuse and violations. This finding speaks to international legal compliance scholarship and supports the idea that states are being sincere about their intentions and not, byinlarge, using actions like derogation to cover abusive behavior.

The rest of the paper proceeds as follows. First, we explore the purpose of derogations in international law, and then specifically during the pandemic. In doing so, we focus on derogations to the International Covenant on Civil and Political Rights (ICCPR)—arguably the most significant and extensive treaty in the international human rights regime. Following this, we introduce our expectations regarding how formally derogating from the treaty (or not) may impact a state's human rights practices during the pandemic. We conclude with the implications of our research note for the study of international law during times of crises.

@@

Derogations in international law authorize states to temporarily suspend their international treaty commitments during times of crises, providing them with flexibility in responding to events. Times of crises can include wars, natural disasters, internal political unrest such as terrorism, and public health emergencies. Derogations afford states the opportunity to adjust to these changing conditions [@HBHelferFariss]. When states derogate from their international legal obligations, the derogation provides vital information to both international and domestic monitoring bodies, interest groups, and advocates about which rights are suspended, for how long, and the reasoning behind these suspensions. This information allows these actors—at least in principle—to challenge measures that are excessive, vague, or outlast the intended time frame of their implementation [@helfer2021].

While legal critical assessments of derogations find importance in their use [@richardson2020], others argue that, "When a state does not bother with derogation provisions in times of emergencies where human rights are suspended, then it is a situation of unbridled state power.” [@burchill2005, 97] An abuse of state power can also potentially lead to the violation of non-derogable rights or rights whose protections cannot be affected by the declaration of a de jure emergency (Greene 2020, 36). These include rights such as right to life, freedom from torture, freedom from slavery, freedom of thought and conscience, among others, that are viewed as essential rights that cannot be suspended even during emergencies. Research finds results that support this notion—there is no meaningful variation in states' respect for derogable or non-derogable rights during officially declared states of emergencies [@richardsclay2012; @neumayer2013]. Even states that officially derogate from their international legal obligations are more likely to violate physical integrity rights, which are non-derogable rights [@neumayer2013; @comstock2019].

These findings suggest that states may not be discerning the significance of derogable v/s. non-derogable rights status during emergencies. This also raises the question of the extent to which this distinction is a valuable one during the pandemic. Most derogations to the ICCPR have been for internal conflict or natural disaster. Is a public health emergency different from other states of emergency? Overall, there is a general expectation that following derogations, we may observe some decline in respect for human rights. However, previous studies have not explored the relationship between derogations and public health emergencies.

@@

A first question of interest before we examine the potential impact of derogations is the nature of states that are more likely to submit them in the first place. Little research has focused on derogations in political science, and research that does exist has mostly focused on the determinants of submitting derogations to the United Nations (UN) rather than the effects of these derogations. This literature tells us that states that are more democratic are more likely to issue derogations and other post-commitment actions such as reservations, understandings, and declarations [@Simmons2009; @comstock2019] and that most states facing an emergency do not, in fact, issue derogations [@HBHelferFariss]. The expectation is that states with strong democratic institutions need to convince domestic audiences of the legality and legitimacy of suspending rights during emergencies. Additionally, democratic states understand international law and international legal mechanisms to be more legitimate and recognize their importance more so than other states. Less democratic states, even though at times may be more likely to violate or evade international human rights law, are less likely to see nuanced pieces of treaty law as important.

@@

To legally derogate, government must first officially declare a state of emergency (Article 4). Derogations must fulfill three criteria: they should be temporary in application, proportional to a crisis' actual threat and non-discriminatory. However, it is worth noting that states cannot derogate from the entirety of the ICCPR. Seven rights enshrined in the ICCPR are non-derogable: the right to life, the prohibition of torture or cruel, inhumane punishment, the prohibition of slavery, the right to recognition as a person before the law, and the protection of freedom of thought, conscience and religion. The international legal community viewed these rights as so fundamental and important that no type of crisis could justify their suspension. Other types of rights are seen as able to be suspended during crises. Several ICCPR Articles provide that public health needs can justify limitations on or derogations from certain rights [@richardson2020, 112]: freedom of movement (Article 12), freedom of expression (Article 19), and the rights to peaceful assembly and freedom of association (Article 22).

Recognizing that public health exigencies would require states to curb civil and political rights, the UN Human Rights Office of the High Commissioner (OHCHR) published guidance stating that any emergency measures must meet specific criteria of legality, necessity, proportionality, and non-discrimination. However, not every country formally derogated—a number of countries implemented emergency measures without formally derogating from ICCPR. Over 100 countries issued emergency declarations while fewer than 30 submitted COVID-19 related derogations [@icnl-covid]. These COVID-19 emergency declarations may have longer term impacts beyond temporary human rights violations [@BrownBrechenmacherCarothers:2020]. Below, we first theorize how derogations led to variation in the implementation of emergency measures during the pandemic, followed by their impact on human rights.

# The impact of derogations during Covid-19

During the pandemic, as @helfer2021 notes, states have had “sharply divergent responses” to international human rights law even when they are parties to the same treaty and confronting similar threats (21).

AUDREY - write expectations about what how derogating and non-derogating states differed in their likelihood of imposing emergency policies - canceling public events, gathering restrictions, closing public transit, movement and international travel

@@

If states determine that emergency measures are justified without regard for international law, the processes for respecting rights during crises and restoring rights at the end of the emergency may not be followed. As such, many states used the pandemic an as excuse to centralize authority and implemented measures that did not meet international legal standards. Many states imposed in discriminatory and abusive policy enforcement--they imposed restrictions on travel or movement by minority or opposition groups. For instance, in Australia, indigenous and migrant communities were disproportionately targeted in the enforcement of movement restrictions [@ICRC2021]. In Bulgaria, checkpoint controls went into effect in neighborhoods largely composed of the Roma community [@bulgaria2020]. In India, the police used a ban on public gatherings as an excuse to get rid of a months-long sit-in protesting the country's new citizenship law seen as discriminatory towards Muslims [@bbc2020].

The pandemic also led to the violation of non-derogable rights, or rights that cannot be suspendended in any circumstances, in many states. For instance, Greek officials intercepted and turned back boats filled with asylum-seekers and also summarily returned asylum-seekers at the land border with Turkey. In doing so, they both violated the non-derogable principle of non-refoulement, as well as Article 7 of ICCPR is the person sent back to another state where they are in danger of ill-treatment or torture [@HathawayStevensLim:2020]. Many other states  changed rules of confinement affecting detainees, which also potentially violated non-derogable rights. For instance, in March 2020, a French government order extended the duration of pre-trial detention, raising the issue of prison overcrowding and leading to inhumane and degrading treatment [@Lebret:2020]. As such, this was not just a violation of a non-derogable right, but also raised questions regarding the proportionality of emergency measures.

We expect that states that did not formally derogate and did not notify the ICCPR of impending rights suspensions and submit derogations were less likely to take the criteria for derogations—necessity, proportionality, non-discrimination and time delimited seriously, and therefore more likely to impose measures that violated these criteria. If states determined that emergency measures are justified without consideration of international law, we expect that there will be more violations.

@@

@@

While some studies on repression, human rights, and the pandemic have focused within single country cases or across a region, we examine a global database to more broadly examine the relationship between treaty derogations, human rights, and state actions during the pandemic. We explore our hypotheses using data on COVID-19 derogations, state actions, health trends, and civil society measures that we collected from several different sources (see @tbl-data-sources), including start and end dates of ICCPR derogations [@UnitedNations:2023], formal government responses from the Oxford Covid-19 Government Response Tracker [@HaleAngristGoldszmidt:2021], indicators of the severity of government responses from the Varieties of Democracy project [@vdem-v12; @vdem-pandem], and COVID case and death counts from the World Health Organization's COVID-19 dashboard [@WHO:2023]. We include data from `r info$n_countries` countries during the first 15 months (or `r info$n_weeks` weeks) of the pandemic, from March 11, 2020 to June 30, 2021. While the Oxford Government Response Tracker provides daily data on a variety of government actions, submitting formal treaty responses to the ICCPR is less likely to be as immediate of an action as implementing domestic emergency actions like stay at home orders. Accordingly, we collapse this daily data into weekly totals, resulting in a panel with `r info$n_rows` country-week observations.

@@

# Conclusion

The Covid-19 pandemic raised numerous concerns that the cris would contribute to violations of democratic standards and human rights in a number of countries and would reinforce the powers of repressive governments worldwide. Therefore, this research note set out to examine state responses to their international human rights treaty obligations, particularly the ICCPR, during the COVID-19 pandemic and any subsequent impact on human rights violations. We focused specifically on the impact of derogations--an international legal means to communicate to both the UN and global community that certain rights will be suspended during times of crisis. Derogations are meant to provide states with flexibility during crises and to encourage them to sign and ratify treaties without concern about the inability to adapt to times of emergency. International legal scholarship has been divided over expectations of derogations during the pandemic. @Lebret:2020 argued that derogations are the best legal mechanism for states to notify the global community of their inability to comply with human rights standards during the COVID-19 pandemic. However, @richardson2020 see derogations as a less clear path for states during COVID-19, describing the actions, their conditions, and human rights as a "tangled morass" (p. 106).

With the caveat that our results are by no means causal—we have not attempted to use any formal strategy to identify an unconfounded causal effect, and our findings are merely descriptive—we find that we find that there is good news for compliance and derogations. Derogators did not systematically use this legal "escape clause" opportunistically—derogators used the process correctly. Most derogations included time limitations, states tended to adhere to the time limits, and there was not an overall abuse of the derogation time to abuse, repress, and discriminate. Though a majority of states did not submit COVID-19 related derogations, the ones that did used derogations as intended by international law. Once the derogation ended, states readjusted to typical compliance once more. States, therefore complied with the intent of derogations and the purpose of the treaty. This is an important finding for international legal behavior and compliance studies. Even in a time of crisis, overall, states considered it important to follow through with the processes and expectations of international human rights law bureaucracy.

We suggest that future research look to different incentives facing state officials during crisis. There may be key differences in enforcement between local level bureaucrats and higher level elites. While higher level diplomats submit derogations and may be more cautious about international law, local level enforcement may not be as concerned about international legal aspects such as time limits and derogable vs. non-derogable rights. This distinction may explain some variation in policy and discrimination practices during the pandemic. We also recommend that future research analyze the impact of other factors that can mitigate the adverse impacts of emergency measures.  In the absence of adequate review from the international community and treaty monitoring bodies during crises, certain domestic-level factors, such as a robust civil society and media, may be necessary to mitigate the adverse effects of emergency measures.
